%\vita
\chapter*{Vita}
\addtocontents{toc}{\vspace{\normalbaselineskip}}
\addcontentsline{toc}{chapter}{\hskip 2.0em VITA}
\begin{singlespace}
    \begin{description}[labelwidth=4cm,leftmargin=4.2cm,itemsep=1em]

        % Your full name bere
        \item[NAME] Alexander Politowicz

        % Put any degrees you have, in the form of: Degree type, Degree Field, Institution, City, State, Year
        \item[EDUCATION] 
            \begin{itemize}[label={},listparindent=0pt,itemindent=0pt,leftmargin=0pt,itemsep=1em,parsep=0pt,topsep=0pt,partopsep=0pt]
            \item Ph.D., Computer Science, University of Illinois Chicago, Chicago, IL, 2026
            \item B.S., Computer Science, University of Wisconsin - Madison, Madison, WI, 2019
            \item B.S., Mechanical Engineering, University of Wisconsin - Madison, Madison, WI, 2019
            \end{itemize}

        % If you have any teaching experience, put it here
        \item[EXPERIENCE]
            \begin{itemize}[label={},listparindent=0pt,itemindent=0pt,leftmargin=0pt,itemsep=1em,parsep=0pt,topsep=0pt,partopsep=0pt]
            \item Graduate Research Assistant, University of Illinois Chicago, Chicago, IL, 2019 - 2026
            Worked on research projects involving Out-of-Distribution Detection for both Image and Text Data, Novelty Detection in Data, Automatic Reward Shaping in Reinforcement Learning, and LLM Fairness.
            \item Graduate Teaching Assistant, University of Illinois Chicago, Chicago, IL, 2019 - 2026
            Taught Systems Programming and Program Design.
            \item Instructor, University of Illinois Chicago, Chicago, IL, 2023
            Taught Program Design as part of the CHANCE program.
            \item Undergraduate Research Assistant, University of Wisconsin - Madison, Madison, WI, 2018 - 2020
            Worked on research projects involving ML modeling of steel embrittlement, prediction of material properties, and uncertainty estimation for materials science data.
            \item Undergraduate Research Assistant, University of Wisconsin - Madison, Madison, WI, 2016 - 2017
            Worked on research projects involving fabrication of hydrophobic oleophilic aerogels.
            \end{itemize}

        \item[PUBLICATIONS]
            \begin{itemize}[label={},listparindent=0pt,itemindent=0pt,leftmargin=0pt,itemsep=1em,parsep=0pt,topsep=0pt,partopsep=0pt]

                % Place each publication you've completed, no matter how trivial,
                % here.  Match the formatting below, to match the expected formatting.
                \item Alexander Politowicz, Sahisnu Mazumder, and Bing Liu. %
                ``Achieving State-of-the-Art Out-of-Distribution Detection Performance with only LLM Prompting'' %
                Under Review at the International Joint Conference on Artificial Intelligence (IJCAI, 2026).

                \item Alexander Politowicz, Sahisnu Mazumder, and Bing Liu. %
                ``Improving Out-of-Distribution Detection using Segmented Images and Cross-View Attention Fusion'' %
                In Proceedings of the Winter Conference on Applications of Computer Vision (WACV, 2026).

                \item Alexander Politowicz, Sahisnu Mazumder, and Bing Liu. %
                ``Safety through Permissibility: Shield Construction for Fast and Safe Reinforcement Learning'' %
                arXiv preprint arXiv:2405.19414.

                \item Alexander Politowicz, Andrew Burks, Yushen Dong, Yu Maw Htwe, Steven Dudek, G Elisabeta Marai, Patrick Belvitch. %
                ``Alveolus analysis: a web browser-based tool to analyze lung intravital microscopy'' %
                BMC Pulmonary Medicine, 2022.
                
                \item Nianzu Ma, Sahisnu Mazumder, Alexander Politowicz, Bing Liu, Eric Robertson, Scott Grigsby.
                ``Detecting and Characterizing Semantic Novelties in Factual Text Involving Named Entities''
                In Proceedings of the 2022 conference on Empirical Methods in Natural Language Processing (EMNLP, 2022).

                \item Glenn Palmer, Siqi Du, Alexander Politowicz, Joshua Paul Emory, Xiyu Yang, Anupraas Gautam, Grishma Gupta, Zhelong Li, Ryan Jacobs, Dane Morgan.
                ``Calibration after bootstrap for accurate uncertainty quantification in regression models''
                npj Computational Materials, 2022.

                \item Yu-chen Liu, Henry Wu, Tam Mayeshiba, Benjamin Afflerbach, Ryan Jacobs, Josh Perry, Jerit George, Josh Cordell, Jinyu Xia, Hao Yuan, Aren Lorenson, Haotian Wu, Matthew Parker, Fenil Doshi, Alexander Politowicz, Linda Xiao, Dane Morgan, Peter Wells, Nathan Almirall, Takuya Yamamoto, G Robert Odette.
                ``Machine learning predictions of irradiation embrittlement in reactor pressure vessel steels''
                npj Computational Materials, 2022.

                \item Nianzu Ma, Alexander Politowicz, Sahisnu Mazumder, Jiahua Chen, Bing Liu, Eric Robertson, Scott Grigsby.
                ``Semantic novelty detection in natural language descriptions''
                In Proceedings of the 2021 conference on Empirical Methods in Natural Language Processing (EMNLP, 2021).

                \item Xiaoyu Sun, Nathaniel J Krakauer, Alexander Politowicz, Wei‐Ting Chen, Qiying Li, Zuoyi Li, Xianjia Shao, Alfred Sunaryo, Mingren Shen, James Wang, Dane Morgan.
                ``Assessing graph‐based deep learning models for predicting flash point''
                Molecular Informatics, 2020.

                % etc...
            \end{itemize}
    \end{description}
\end{singlespace}